Neural Power Units

### 2020-06-05 papers/main.tex
pevnak fixes

\title{Neural Power Units}
\author{
  Niklas Heim,
  Tom\'a\v s Pevn\'y,
  V\'aclav \v Sm\'idl\\
  Artificial Intelligence Center\\
  Czech Technical University\\
  Prague, CZ 120 00\\
  \texttt{\{niklas.heim, vasek.smidl, tomas.pevny\}@aic.fel.cvut.cz}\\
}

@@

Including Neural Arithmetic Units in standard neural networks, promises to
significantly increase their extrapolation capabilities due to their inductive
bias towards numerical computation. This is especially true for tasks in which
the data generating process contains mathematical relationships.  Additionally,
they promise to reduce the number of parameters that are needed for a given
task, and can therefore drastically improve the explainability of the model.
We demonstrate how a \emph{Neural Ordinary Differential Equation} (NODE,
\cite{chen_neural_2019}) composed of small neural arithmetic units can
outperform a much larger NODE built from dense layers (Sec.~\ref{sub:fractional_sir_model}).
In fact, our new unit can be used to directly read out the correct generating ODE
from the fitted model.  This is in line with recent efforts to build
\emph{interpretable} models instead of \emph{explaining} black-box models
\citep{rudin_stop_2019}, like conventional neural networks.

The currently available arithmetic units all have different strengths and
weaknesses, but none of them solve simple arithmetic completely.
The \emph{Neural Arithmetic Logic Unit} (NALU) by \cite{trask_neural_2018},
chronologically, was the first arithmetic unit. It can solve
addition ($+$, including subtraction), multiplication ($\times$), and division
($\div$), but is limited to positive inputs. The convergence of the
NALU is quite fragile due to an internal gating mechanism between addition and
multiplication paths as well as the use of a logarithm which is problematic for
small inputs.  Recently, \cite{schlor_inalu_2020} introduced the \emph{improved NALU} (iNALU,
to fix the NALU's shortcomings. However, it significantly increases its complexity, and we
observe only a slight improvement in performance.
\cite{madsen_neural_2020} solve ($+,\times$) with two new units: the
\emph{Neural Addition Unit} (NAU), and the \emph{Neural Multiplication Unit}
(NMU). Instead of gating between addition and multiplication paths, they are
separate units that can be stacked. They can work with the full range of real
numbers, converge much more reliably, but cannot represent
division.

@@

The NaiveNPU has difficulties to converge on large scale tasks, and to reach
sparse results in cases where the input to a given row is small. We
demonstrate this on a toy example of learning the identity on one of two inputs
(neglecting the second one):
\begin{align}
  \nonumber
  \mathcal{L} &= |\text{m}(x_1,x_2) - x_1|, \quad \text{where}
  & \text{m} = \npu(2,1), \quad
  &x_1 \sim \mathcal U(0,2), \quad x_2 \sim \mathcal U(0,0.05).
\end{align}
The left plot in Fig.~\ref{fig:id_loss} depicts the gradient norm of NPU and
NaiveNPU for a batch of two-dimensional inputs. One input is small and
irrelevant. Even in this simple example, gradient of the NaiveNPU is close to
zero in large parts of the parameter space This can be explained as follows.
One row of NaiveNPU weights effectively raises each input to a power and
multiplies them: $x_1^{w_1} x_2^{w_2} \dots x_n^{w_n}$. If a single input $x_i$
is constantly close to zero (i.e. irrelevant), the whole row will be zero, no
matter what its weights are and the gradient information on all other weights
is lost.  Therefore, we introduce a gate on the input of the NPU that can turn
irrelevant inputs into 1s.
A diagram of the NPU is shown in Fig.~\ref{fig:gatednpu_diagram}.
\begin{definition}[{\bf NPU}]
  The NPU extends the NaiveNPU by the relevance gate $\bm g$ on the input $\bm x$.
   \begin{gather}
    \label{eq:gatednpu_def}
    \bm z = \exp(\Wre \log\bm r - \pi\Wim\bm k) 
          \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where } \\
    \bm r = \bm{\hat g} \odot |\bm x| + (1-\bm{\hat g}),
    \quad
    k_i = \begin{cases}
       0  & x_i \leq 0 \\
      \hat g_i & x_i > 0
    \end{cases},
    \quad
    \hat g_i = \min(\max(g_i,0),1)
  \end{gather}
\end{definition}
\begin{figure}
  \centering
  \begin{minipage}{0.55\textwidth}
    \includegraphics[width=\textwidth]{npu_gatednpu_id_grad.pdf}
  \end{minipage}
  \begin{minipage}{0.42\textwidth}
  \end{minipage}
  \caption{Norm of the gradient of NaiveNPU and NPU for the task
  of learning the identity on $x_1$. Inputs and
  loss are defined on the right, gradient surfaces on the left
  (black areas are beyond the color scale). The correct solution is
  $w_1=1$ and $w_2=0$.  The NaiveNPU has a large zero gradient region for
  $w_2>0.75$, while the NPU's surface is much more informative. The gates
  for central plot are fixed at $g_1=g_2=0.5$ which corresponds to the initial
  gate parameters. During training they will adjust as needed, in this case to
  $g_1=1$ and $g_2=0$. $\bm W_i$ is set to zero in all plots.
  }%
  \label{fig:id_loss}
\end{figure}
The central plot of Fig.~\ref{fig:id_loss} shows the gradient of the NPU
on the identity task with its initial gate setting of $g_1=g_2=0.5$. The
large zero-gradient region of the NaiveNPU is gone. The last plot shows the
same loss for $g_1=1$ and $g_2=0$, which corresponds to the correct gates at
the end of NPU training. The gradient is independent of $w_2$, which
means that it can easily be pruned by a simple regularization such as L1.  In
Sec.~\ref{sub:large_scale_arithmetic_task} we show how important the relevance
gating mechanism is for the convergence and sparsity of large models.  Sparsity
is especially important in order to use the NPU as an interpretable model.

@@

If an ODE model is composed of dense layers, its direct interpretation is
problematic and has to be performed retrospectively. The class of models based on
SINDy is interpretable by design, however it can only provide explanation
within a linear combination of predefined set of basis functions. Thus, it
cannot learn models with unknown fractional powers. We will demonstrate that the NPU
is capable to do so.
An example of an ODE that contains powers is a modification of the classical
epidemiological SIR model \citep{kermack1927contribution} to fractional powers
(fSIR, \cite{taghvaei_fractional_2020}), which was shown to be a beneficial
modification for modelling the COVID-19 outbreak.  The classical SIR model is
built from three variables: $S$ (susceptible), $I$ (infectious), and $R$
(recovered/removed).  Arguably the most important part of the model is the
transmission rate $r$, which is typically taken to be proportional to the
product of $S$ and $I$.  \citet{taghvaei_fractional_2020} argue that,
especially in the initial phase of an epidemic, the boundary areas of infected
and susceptible cells scale with a fractional power, which leads to
Eq.~\ref{eq:fracsir_r}:
\begin{align}
  \label{eq:fracsir}
  \frac{dS}{dt} &= -r(t) + \eta R(t), &
  \frac{dI}{dt} &=  r(t) - \alpha I(t), &
  \frac{dR}{dt} &= \alpha I(t) - \eta R(t),\\
  \label{eq:fracsir_r}
  & & r(t) &= \beta I(t)^\gamma S(t)^\kappa,
\end{align}

@@

The RealNPU has problems to converge in the majority of cases.
However, there are a few models in the bottom left that

>>>>

\title{Neural Power Units}
\author{Anonymous Authors}

@@

Including Neural Arithmetic Units in standard neural networks, promises to
significantly increase their extrapolation capabilities due to their inductive
bias towards numerical computation. This is especially important for tasks in which
the data generating process contains mathematical relationships.  They also
promise to reduce the number of parameters needed for a given
task, which can improve the explainability of the model.
We demonstrate this in a \emph{Neural Ordinary Differential Equation} (NODE,
\cite{chen_neural_2019}), where a handful of neural arithmetic units can
outperform a much bigger network built from dense layers (Sec.~\ref{sub:fractional_sir_model}).
Moreover, our new unit can be used to directly read out the correct generating ODE
from the fitted model.  This is in line with recent efforts to build
\emph{interpretable} models instead of \emph{explaining} black-box models
\citep{rudin_stop_2019}, like conventional neural networks.

The currently available arithmetic units all have different strengths and
weaknesses, but none of them solve simple arithmetic completely.
The \emph{Neural Arithmetic Logic Unit} (NALU) by \cite{trask_neural_2018},
chronologically, was the first arithmetic unit. It can solve
addition ($+$, including subtraction), multiplication ($\times$), and division
($\div$), but is limited to positive inputs. The convergence of the
NALU is quite fragile due to an internal gating mechanism between addition and
multiplication paths as well as the use of a logarithm which is problematic for
small inputs.  Recently, \cite{schlor_inalu_2020} introduced the \emph{improved NALU} (iNALU,
to fix the NALU's shortcomings. It significantly increases its complexity, and we
observe only a slight improvement in performance.
\cite{madsen_neural_2020} solve ($+,\times$) with two new units: the
\emph{Neural Addition Unit} (NAU), and the \emph{Neural Multiplication Unit}
(NMU). Instead of gating between addition and multiplication paths, they are
separate units that can be stacked. They can work with the full range of real
numbers, converge much more reliably, but cannot represent
division.

@@

The NaiveNPU has difficulties to converge on large scale tasks, and to reach
sparse results in cases where the input to a given row is small. We
demonstrate this on a toy example of learning the identity on one of two inputs and neglecting the second one, $f(x_1,x_2) = x_1$:
\begin{align}
  \nonumber
  \mathcal{L} &= |\text{m}(x_1,x_2) - x_1|, \quad \text{where}
  & \text{m} = \npu(2,1), \quad
  &x_1 \sim \mathcal U(0,2), \quad x_2 \sim \mathcal U(0,0.05).
\end{align}
The left plot in Fig.~\ref{fig:id_loss} depicts the gradient norm of NPU and
NaiveNPU for a batch of two-dimensional inputs. One input is small and
irrelevant. Even in this simple example, gradient of the NaiveNPU is close to
zero in large parts of the parameter space This can be explained as follows.
One row of NaiveNPU weights effectively raises each input to a power and
multiplies them: $x_1^{w_1} x_2^{w_2} \dots x_n^{w_n}$. If a single input $x_i$
is constantly close to zero (i.e. irrelevant), the whole row will be zero, no
matter what its weights are and the gradient information on all other weights
is lost.  Therefore, we introduce a gate on the input of the NPU that can turn
irrelevant inputs into 1s.
A diagram of the NPU is shown in Fig.~\ref{fig:gatednpu_diagram}.
\begin{definition}[{\bf NPU}]
  The NPU extends the NaiveNPU by the relevance gate $\bm g$ on the input $\bm x$.
   \begin{gather}
    \label{eq:gatednpu_def}
    \bm z = \exp(\Wre \log\bm r - \pi\Wim\bm k) 
          \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where } \\
    \bm r = \bm{\hat g} \odot |\bm x| + (1-\bm{\hat g}),
    \quad
    k_i = \begin{cases}
       0  & x_i \leq 0 \\
      \hat g_i & x_i > 0
    \end{cases},
    \quad
    \hat g_i = \min(\max(g_i,0),1)
  \end{gather}
\end{definition}
\begin{figure}
  \centering
  \begin{minipage}{0.55\textwidth}
    \includegraphics[width=\textwidth]{npu_gatednpu_id_grad.pdf}
  \end{minipage}
  \begin{minipage}{0.42\textwidth}
  \end{minipage}
  \caption{Norm of the gradient of NaiveNPU and NPU for the task
  of learning the identity on $x_1$. Inputs and
  loss are defined on the right, gradient surfaces on the left
  (black areas are beyond the color scale). The correct solution is
  $w_1=1$ and $w_2=0$.  The NaiveNPU has a large zero gradient region for
  $w_2>0.75$, while the NPU's surface is much more informative. The gates
  for central plot are fixed at $g_1=g_2=0.5$ which corresponds to the initial
  gate parameters. During training they will adjust as needed, in this case to
  $g_1=1$ and $g_2=0$. $\bm W_i$ is set to zero in all plots.
  }%
  \label{fig:id_loss}
\end{figure}
The central plot of Fig.~\ref{fig:id_loss} shows the gradient of the NPU
on the identity task with its initial gate setting of $g_1=g_2=0.5$. The
large zero-gradient region of the NaiveNPU is gone. The last plot shows the
same loss for $g_1=1$ and $g_2=0$, which corresponds to the correct gates at
the end of NPU training. The gradient is independent of $w_2$, which
means that it can easily be pruned by a simple regularization such as L1.  In
Sec.~\ref{sub:large_scale_arithmetic_task} we show how important the relevance
gating mechanism is for the convergence and sparsity of large models.  Sparsity
is especially important in order to use the NPU as an interpretable model.

@@

If an ODE model is composed of dense layers, its direct interpretation is
problematic and has to be performed retrospectively. The class of models based on
SINDy is interpretable by design, however it can only provide explanation
within a linear combination of predefined set of basis functions. Thus, it
cannot learn models with unknown fractional powers. We will demonstrate that the NPU
is capable of doing so.
An example of an ODE that contains powers is a modification of the classical
epidemiological SIR model \citep{kermack1927contribution} to fractional powers
(fSIR, \cite{taghvaei_fractional_2020}), which was shown to be a beneficial
modification for modelling the COVID-19 outbreak.  The classical SIR model is
built from three variables: $S$ (susceptible), $I$ (infectious), and $R$
(recovered/removed).  Arguably the most important part of the model is the
transmission rate $r$, which is typically taken to be proportional to the
product of $S$ and $I$.  \citet{taghvaei_fractional_2020} argue that,
especially in the initial phase of an epidemic, the boundary areas of infected
and susceptible cells scale with a fractional power, which leads to
Eq.~\ref{eq:fracsir_r}:
\begin{align}
  \label{eq:fracsir}
  \frac{dS}{dt} &= -r(t) + \eta R(t), &
  \frac{dI}{dt} &=  r(t) - \alpha I(t), &
  \frac{dR}{dt} &= \alpha I(t) - \eta R(t),\\
  \label{eq:fracsir_r}
  & & r(t) &= \beta I(t)^\gamma S(t)^\kappa,
\end{align}

@@

The RealNPU has problems to converge in the majority of cases, however, 
there are a few models in the bottom left that
